Resolvable Ambiguity

### 2019-07-08 paper/paper.tex
Almost finished first draft of Orc case study

\usepackage{listings}
\lstset{
  basicstyle=\ttfamily,
  basewidth={.5em,.5em},
}
\lstdefinelanguage{syncon}
  {morekeywords={syncon,infix,postfix,prefix,left,right,comment,precedence,forbid,except,type,builtin,rec,grouping},
   sensitive=true,
   morecomment=[l]{//},
   morecomment=[s]{/*}{*/},
   morecomment=[s][<.][.>],
   morestring=[b]",
  }
\newcommand{\ocaml}{\lstinline[language={[objective]caml}]}
\newcommand{\syncon}{\lstinline[language=syncon]}

@@

\begin{document}

@@

\begin{itemize}
\item A formal definition of \emph{resolvable ambiguity} in terms of languages where words have associated interpretations, as well as formulations of the \emph{static} and \emph{dynamic} resolvability problems, i.e., determine if a grammar (resp. word) is resolvably ambiguous (Section~\ref{sec:resolvable-definition}).
\item A syntax definition formalism based on EBNF, where ambiguities can be resolved with grouping parentheses (Section~\ref{sec:parse-time-disambiguation}). For this formalism we also provide:
  \begin{itemize}
  \item A formulation of three versions of the static resolvability problem of increasing difficulty, and a decidable algorithm that is sound and complete for the first version, and sound for the second (Section~\ref{sec:static}).
  \item A decidable algorithm that solves the dynamic resolvability problem for languages that have only well-balanced parentheses (Section~\ref{sec:dynamic}).
  \end{itemize}
\item Two case studies of programming languages and the errors produced when we allow (resolvable) ambiguity:
  \begin{itemize}
  \item A large subset of OCaml where some ambiguities have been reintroduced (Section~\ref{sec:evaluation-ocaml}).
  \item A domain-specific language for cyber-physical systems (Section~\ref{sec:evaluation-cyphym}).
  \end{itemize}
  % Briefly say something about the tool, that it lies ``on top of'' what's presented herein
\end{itemize}

@@

We have implemented a tool and a small DSL for syntax definition that generates a language definition in the style of Section~\ref{sec:parse-time-disambiguation}, which the tool then uses to construct a parser and do a dynamic resolvability analysis. The tool is implemented in Haskell and uses an off-the-shelf implementation\footnote{\url{http://hackage.haskell.org/package/Earley}} of the Earley parsing algorithm \cite{earleyEfficientContextfreeParsing1970}. The purpose of these case studies is to showcase possible use cases of resolvable ambiguity in the context of a pre-existing general purpose programming language (OCaml, Section~\ref{sec:evaluation-ocaml}), a pre-existing DSL (ORC, Section~\ref{sec:evaluation-orc}), and a new DSL (Cyphym, Section~\ref{sec:evaluation-cyphym}).

The syntax definition DSL is a continuation of the ``syncons'' introduced by \citet{palmkvistCreatingDomainSpecificLanguages2019}, wherein each production is defined separately from each other (one \syncon{syncon} each), and marks are primarily introduced through precedence lists. For example, the running example used in Section~\ref{sec:parse-time-disambiguation} (Fig.~\ref{fig:running-example-definition} on page~\pageref{fig:running-example-definition}) can be written as follows:

\begin{tabular}{ll}
\begin{lstlisting}[language=syncon,boxpos=t]
type Exp
grouping "(" Exp ")"
precedence {
  mul; // higher in precedence list
  add; // means higher precedence
}
\end{lstlisting}
&
\begin{lstlisting}[language=syncon,boxpos=t]
syncon list: Exp =
  "[" (head:Exp (";" tail:Exp))? "]"
infix add: Exp = "+"
infix mul: Exp = "*"
syncon literal: Exp = n:Integer
\end{lstlisting}
\end{tabular}

@@

>>>>

\usepackage{listings}
\lstset{
  basicstyle=\ttfamily,
  basewidth={.5em,.5em},
}
\lstdefinelanguage{syncon}
  {morekeywords={syncon,infix,postfix,prefix,left,right,comment,precedence,forbid,except,type,builtin,rec,grouping,token},
   sensitive=true,
   morecomment=[l]{//},
   morecomment=[s]{/*}{*/},
   morecomment=[s][<.][.>],
   morestring=[b]",
  }
\newcommand{\ocaml}{\lstinline[language={[objective]caml}]}
\newcommand{\syncon}{\lstinline[language=syncon]}

@@

\synctex=1
\begin{document}

@@

\begin{itemize}
\item A formal definition of \emph{resolvable ambiguity} in terms of languages where words have associated interpretations, as well as formulations of the \emph{static} and \emph{dynamic} resolvability problems, i.e., determine if a grammar (resp. word) is resolvably ambiguous (Section~\ref{sec:resolvable-definition}).
\item A syntax definition formalism based on EBNF, where ambiguities can be resolved with grouping parentheses (Section~\ref{sec:parse-time-disambiguation}). For this formalism we also provide:
  \begin{itemize}
  \item A formulation of three versions of the static resolvability problem of increasing difficulty, and a decidable algorithm that is sound and complete for the first version, and sound for the second (Section~\ref{sec:static}).
  \item A decidable algorithm that solves the dynamic resolvability problem for languages that have only well-balanced parentheses (Section~\ref{sec:dynamic}).
  \end{itemize}
\item Three case studies of programming languages and the errors produced when we allow (resolvable) ambiguity:
  \begin{itemize}
  \item A large subset of OCaml where some ambiguities have been reintroduced (Section~\ref{sec:evaluation-ocaml}).
  \item A domain-specific language for orchestrating parallel computations (Section~\ref{sec:evaluation-orc}).
  \item A domain-specific language for cyber-physical systems (Section~\ref{sec:evaluation-cyphym}).
  \end{itemize}
  % Briefly say something about the tool, that it lies ``on top of'' what's presented herein
\end{itemize}

@@

We have implemented a tool and a small DSL for syntax definition
that generates a language definition in the style of
Section~\ref{sec:parse-time-disambiguation}, which the tool then
uses to construct a parser and do a dynamic resolvability
analysis. The tool is implemented in Haskell and uses an
off-the-shelf
implementation\footnote{\url{http://hackage.haskell.org/package/Earley}}
of the Earley parsing algorithm
\cite{earleyEfficientContextfreeParsing1970}. The purpose of these
case studies is to showcase possible use cases of resolvable
ambiguity in the context of a pre-existing general purpose
programming language (OCaml, Section~\ref{sec:evaluation-ocaml}),
a pre-existing DSL (Orc, Section~\ref{sec:evaluation-orc}), and a
new DSL that is under development (Cyphym, Section~\ref{sec:evaluation-cyphym}).

The syntax definition DSL builds on the the ``syncons''
introduced by
\citet{palmkvistCreatingDomainSpecificLanguages2019}, wherein each
production is defined separately from each other (one
\syncon{syncon} each), and marks are primarily introduced through
precedence lists. For example, the running example used in
Section~\ref{sec:parse-time-disambiguation}
(Fig.~\ref{fig:running-example-definition} on
page~\pageref{fig:running-example-definition}) can be written as
follows:

\begin{tabular}{ll}
\begin{lstlisting}[language=syncon,boxpos=t]
type Exp
grouping "(" Exp ")"
precedence {
  mul; // higher in precedence list
  add; // means higher precedence
}
\end{lstlisting}
&
\begin{lstlisting}[language=syncon,boxpos=t]
token Integer = "[0-9]+"
syncon list: Exp =
  "[" (head:Exp (";" tail:Exp))? "]"
infix add: Exp = "+"
infix mul: Exp = "*"
syncon literal: Exp = n:Integer
\end{lstlisting}
\end{tabular}\smallskip

@@

\subsection{Orc} \label{sec:evaluation-orc}

Orc~\cite{kitchinOrc2009} is a functional programming language
which includes a number of special-purpose combinators for
coordinating concurrent workflows. In this section, we consider
the use case of adding these concurrency combinators to an already
defined programming language.

In Orc, every expression can ``publish'' zero or more values. Orc
defines four combinators for orchestrating these published values:
the parallel (\ocaml{|}), sequential (\ocaml{>}$x$\ocaml{>}),
pruning (\ocaml{<}$x$\ocaml{<}), and ``otherwise'' (\ocaml{;})
combinators. The expression $e_1$ \ocaml{|} $e_2$ runs $e_1$ and
$e_2$ in parallel, publishing any value published by either of
them. The expression $e_1$ \ocaml{>}$x$\ocaml{>} $e_2$ binds each
value published by $e_1$ to $x$ and executes $e_2$, building a
concurrent pipeline. The expression $e_1$ \ocaml{<}$x$\ocaml{<}
$e_2$ binds the \emph{first} value published by $e_2$ to $x$ and
executes $e_1$. Finally, $e_1$\ocaml{;}$e_2$ runs $e_2$ only if
running $e_1$ does not publish any values.

The syncon definition of the Orc combinators is very simple (the
precedence rules follow the original
definition~\cite{kitchinOrc2009}):

\begin{tabular}{ll}
\begin{lstlisting}[language=syncon,boxpos=t]
type Exp
grouping "(" Exp ")"
precedence {
  seq; par; prune; otherwise;
}
\end{lstlisting}
&
\begin{lstlisting}[language=syncon,boxpos=t]
token Ident = "[[:lower:]][[:word:]]*"
infix par:Exp = "|"
infix seq:Exp = ">" x:Ident ">"
infix prune:Exp = "<" x:Ident "<"
infix otherwise:Exp = ";"
\end{lstlisting}
\end{tabular}\smallskip

\noindent
Na\"{i}vely composing this definition with another, separately
defined language is likely to introduce ambiguities. For example,
if that language also contains infix operators, the precedence
between these and the Orc combinators will be undefined. With
support for resolvable ambiguity however, we can allow this
ambiguity and let programmers disambiguate expressions using
parentheses. After composing the above definition with a simple
language supporting addition, variables and function calls
(omitted for brevity), parsing the expression
%
``\ocaml{1 + 2 >x> f(x)}''
%
results in the following error message from our tool:

\begin{lstlisting}[language={[objective]caml}]
Ambiguity error with 2 alternatives:
  ( 1 + 2 ) > x > f ( x )
  1 + ( 2 > x > f ( x ) )
\end{lstlisting}

\noindent
In this case, it is likely that the preferred semantics is to have
all operators in the base language bind tighter than the Orc
combinators, and these precedence rules can be added after the
composition without changing the original definitions.

% TODO: Clashes involving ">" or ";"?
